Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-03-07 Delaneyetal2024.tex

change subglacial hydraulics to subglacial model outputs??

\begin{keypoints}
\item Short term water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge, as opposed to subglacial sediment transport capacity in glacierized catchments. \mauro{I don't understand this sentence.}
\end{keypoints}

\begin{abstract}  Sediment transport capacity in subglacial and subaerial channels depends on  water velocity and channel width, which together control the total shear stress exerted on the channel bottom.
  In subaerial channels water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to slow-evolving channel geometry.
  Parameterizations of these subglacial and subaerial hydraulics are applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results show that high subglacial sediment transport rates can occur across a wide range of water discharges.
  Slowly evolving channel width accommodates some variability in subglacial channels. \mauro{not sure what that sentence should do}
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulic parameters.
  This implies that other factors like sediment access or proglacial processes might be at play that affect sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

@@

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau_t \propto w\,v^2.
  \end{equation}
\end{linenomath*}
\mauro{I removed the width from here as I found it odd to introduce that here.}
%
Following mass conservation, the velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area. \mauro{I find it odd to have formulas in the Intro}

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a commensurate increase in water velocity.
%Thus,  fulfilling Equation~\ref {eq:v}, the higher discharge is accommodated by both an increase in wetted area and flow velocity. \mauro{not needed}
As a result, the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by melting due to frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
%It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w$ evolves as well.

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_g$ remain unchanged, while the subaerial channel widths $w_f$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
      %Note that in the subaerial channel parameterization, channels have rectangular shapes with their width much larger than their depth (Section~\ref{sect:fluv}).
\mauro{Needs renaming $w_{cg/f}$ to $w_{g/f}$. Then, I find the downward arrow confusing.  Maybe just leave away? Also, I would write $\tau_f$ and $\tau_g$, it's ok to not have the width integrated tau here (and maybe we remove $\tau_{tf}$ from the paper anyway...}
    }
    \label{fig:cartoon}
  \end{figure}
% \ian{Come back to Daniel's comments}

Consequently, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Due to the absence of continuous measurements of subglacial and subaerial water velocity and channel morphology, numerical parameterizations of subglacial and subaerial hydraulics are utilized. \mauro{Aren't those things always parameterised, irrespective of whether there are measurements or not?}
Parameterizations are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
Our parameterizations' lumped nature isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
Furthermore, outputs are compared to sediment discharge data from these two catchments.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport from glacierized catchments.
\mauro{As stated in the general comments.  I'd rather use ``model'' than ``parameterization'' as the latter is, for me, something which is used in a model and not a model itself.}

@@

Water and suspended sediment data are leveraged from Alpine and ice sheet settings.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data were interpolated using a linear spline.

\section{Methods}
\label{sect:meth}
The parameterizations below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both parameterizations calculate   water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

\subsection{Subglacial channel  parameterization}
\label{sect:sub_mode}

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel parameterization takes into account the channel geometry and the velocity of the flowing water.
To do this, a lumped hydraulics model \cite{clarke1996} modified from the one presented by \citeA{werder2010b} is used.

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to melt due to frictional heating from water flow and to creep closure of the ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
Thus, it is assumed that the changing hydraulic head at the top of the glacier is negligible in the channel size evolution \mauro{I don't understand this sentence}.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of the subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure expressed in meter water equivalent, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_g^2}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the length over which head change occurs, and $v_g=Q/S_g$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
%
The shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity.
%
The width-integrated shear stress is represented as
\begin{equation}
  \label{eq:tautg}
  \tau_{tg}=w_g\,\tau_g,
\end{equation}
where  channel floor width $w_g$ is
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
\mauro{If we get rid of $\tau_{tf}$, above needs a bit of massaging.  I'd then move the $w_g$ eq to just after the $S_g$ eq.}

\subsection{Subaerial channel  parameterization}
\label{sect:fluv}

% To parameterize the shear stress of water flowing across sediments in the subaerial channel, the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels \cite{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
%
where $k$ is a constant and $\alpha=\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_f\,\tau_f.
  \end{equation}

  Note that the subaerial channel model \mauro{or ``parameterization''} is a purely algebraic model, whereas the subglacial model comprises a differential equation for the evolution of $S_g$ and thus will have a history dependence on the discharge.

@@

\mauro{I would call the ``cases'' ``scenarios'' instead.  The word ``scenario'' does not appear in the manuscript so far, so it would be unambiguous.  Also, ``cases'' has the disadvantage that ``In both cases,'' is a saying, and thus it is confusing.  I implemented this change.}

The parameterizations above are applied to hydrological records from the Fieschergletscher (scenario \alpine) and the Leverett Glacier (scenario \icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%
However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{2}$, i.e. a segment of a quarter-circle.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
\mauro{Those $f$ reported in the table are super large!  Maybe an issue down the line...}
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ \mauro{Why do you report a value here but most are hidden in the table?  Maybe just ref the table?}.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameter combinations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$). \mauro{So how many runs are actually run?  Not clear.}
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outwith $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outwith $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013} \mauro{Awkward sentence.  (also note that ``cull'' means to remove, thus I changed ``between'' to ``outwith''.}.
The variability of output is represented by the  standard deviation of that output over the course of the season \mauro{This sentence needs context.  What is variability used for?}.

Lastly, outputs of $w\tau$ are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing, thus making our interpretation independent of the exact details of the sediment transport formula.
Outputs and observations, which have a $15$\,\unit{min} resolution, are aggregated across time intervals of up to $15$\,\unit{days} by averaging values over a symmetric window of corresponding size.

@@

I.D. designed the study, developed, and implemented the model and experiments, and wrote the manuscript.
A.J.T. helped interpret data from Leverett Glacier and provided guidance on experiment design.
M.A.W. provided data from Fieschergletscher and added the relationship between shear stress and water discharge (see Sections~\ref{sect:dis_qsc} and \ref{sect:scaling}).
D.F. provided data from Fieschergletscher.
All authors provided  key inputs in writing and editing the manuscript.

>>>>

%\ian{only glaciers with subglacial channels... }

\begin{keypoints}
\item Short term water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Results suggest that significant sediment transport capacity can persist across a range of water discharge conditions, and timing of peak events occurs under different hydrological conditions.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which vary with the  water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to slow-evolving channel geometry.
  Hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to models of subglacial and subaerial hydraulics, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates can occur across a wide range of water discharges.
  However, slowly evolving channel width accommodates some variability in subglacial channels, compared to subaerial channels.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydrological forcings and assumptions of channel type.
\end{abstract}

\section*{Plain Language Summary} % 200 words
Sediment export from glaciers impacts downstream ecosystems and communities and it has been documented to change in response to climate warming and glacier retreat.
In these catchments, sediment export depends in part on the sediment transport capacity of water flowing below the glacier, or the amount of sediment that could be transported given the abundance of sediment.
This capacity largely depends on the width of the channel and the water's velocity. 
Unlike subaerial channels, like rivers, where the size of the river changes rapidly with respect to water discharge, however, subglacial channels are pressurized by the ice above.
This pressurized flow means that channel size evolves slowly compared to the variations in water discharge, therefore variations in water discharge are largely accommodated by variations in velocity.
Models of these processes show that sediment transport capacity is highly variable in subglacial channels, and a wide range of water discharges can result in comparable sediment transport capacity.
In subaerial channels, by contrast, sediment transport capacity is less variable and depends on a single water discharge quantity.
The channel's different response to water discharge may impact the interpretation of sediment transport records and responds to the type of hydrological forcing applied.

@@

Changes in glacier dynamics and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

@@

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
%
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a commensurate increase in water velocity.
As a result, the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by melting due to frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_g$ remain unchanged, while the subaerial channel widths $w_f$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
    }
    \label{fig:cartoon}
  \end{figure}

Consequently, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify subglacial and subaerial water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport, and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is leveraged from Alpine and ice sheet settings collected downstream of the glacier, assuming that no water storage occurred in the proglacial area.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge was collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge was collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models calculate   water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

\subsection{Subglacial channel  model}
\label{sect:sub_mode}

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel model accounts for the channel geometry and the velocity of the flowing water.
To do this, a lumped hydraulics model \cite{clarke1996} modified from the one presented by \citeA{werder2010b} is used.

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to melt due to frictional heating from water flow and to creep closure of the ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of the subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure expressed in meter water equivalent, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_g^2}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the length over which head change occurs, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
%
The shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau_g}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity.
%
The width of the channel floor $w_g$ is represented as
\begin{equation}
  \label{eq:dh2wc}
  w_g = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}
%
This value is used to establish the integrated shear stress across the channel, $w_g\tau_g$.

\subsection{Subaerial channel  model}
\label{sect:fluv}

% To parameterize the shear stress of water flowing across sediments in the subaerial channel, the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_f^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_f} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_f$ is the friction factor for subaerial channels \cite{tucker1997}.
Channel width $w_f$ is
\begin{equation}
  \label{eq:wcf}
  w_f = k \, Q^{\alpha},
\end{equation}
%
where $k$ is a constant and $\alpha=\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau_g} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_f\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is $w_f\tau_f$.

Note that this subaerial channel model is purely algebraic, whereas the subglacial model comprises a differential equation for the evolution of $S_g$.
Thus the models have a history dependence on the discharge $Q_w$.

@@

The models above are applied to hydrological records from the Fieschergletscher (scenario \alpine) and the Leverett Glacier (scenario \icesheet).
The subglacial and subaerial models represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%
However, both scenarios share a discharge with a similar, relative seasonal amplitude.

For each glacier, the models are first applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$, i.e. a segment of a quarter-circle, (Table~\ref{table:vpm}).
Friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities occur for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

A second experiment aims to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range channel shapes, friction factors and water discharge variability. 
To accomplish this, the water discharge records are aggregated across a select time period in the past ranging from $15$\,\unit{min} up to $15$\,\unit{days}, and the model is run with random  parameter combinations of channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_f$).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The routine runs until $100$ different parameter combinations are accepted from each water discharge aggregation period.
The variability of an output, a key difference between the subaerial and subglacial systems, is represented by  its  standard deviation over the course of the season.

The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity in the text these values together are referred to as ``model outputs''.
Lastly, outputs  are compared with the water discharge from glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing, thus making our interpretation independent of the exact details of the sediment transport formula.

@@

I.D. designed the study, developed, and implemented the model and experiments, and wrote the manuscript.
A.J.T. helped interpret data from Leverett Glacier and provided guidance on experiment design.
M.A.W. provided data from Fieschergletscher and added the relationship between shear stress and water discharge (see \ref{sect:scaling}).
D.F. provided data from Fieschergletscher.
All authors provided  key inputs in writing and editing the manuscript.
